Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-12-05 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, determine the cycles of carbon, water and nutrient, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015].
In this context, the maintenance of tropical forest community structure, composition and functioning is increasingly threatened [@Anderson-Teixeira2013].

Tropical forests are subject to natural disturbance events, specifically tree fall gaps, determining and maintaining the characteristics of ecosystems [@Schnitzer2001]. 
Disturbances change both the biotic and abiotic environment, through modifications of the fluxes of light, heat and water [@Goulamoussene2017], and of the interactions among species such as competition [@Chesson2000].
Understand the drivers of community response to disturbance and the underlying ecological processes is then crucial to anticipate the fate of tropical forests in the changing context [@Chazdon2003a].
Community response has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into the recovery of ecosystem processes and services [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that determined succession, mainly driven by the recruitment of pioneer and heliophilous species, shape post-disturbance trajectories [@Carreno-Rocabado2012;@Poorter2016].
Community trajectories would then rely on deterministic processes and would be predictable according to the disturbance intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate from predictions.
Besides, although the deterministic ecological processes at stake are identified, their relative importance compared to stochasticity remains debated [@Norden2015].
This debate translates into the validation and the extent of the intermediate disturbance hypothesis (IDH).
The IDH assumes the predominance of deterministic processes, and predicts a maximum species diversity after moderate disturbance intensity [@Molino2001;@Kariuki2006a].
Validations of the IDH remain scarce in the long-term and often rely on the analysis of taxonomic richness [@Molino2001]. 
Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. 
More ecological-meaningful analysis would couple richness with species evenness, which reveals the changes in the species abundance distribution, and (ii)species composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

@@

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) examined community recovery time.

>>>>

Tropical forests cover large areas worldwide and represent crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, determine the cycles of carbon, water and nutrient, and ensure cultural and human well-being. 
However, the growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015].
This context specifically threatens the maintenance of community structure, composition and functioning [@Anderson-Teixeira2013].
In tropical forests indeed, tree communities are shaped and maintained by a natural disturbance regime, specifically tree fall gaps[@Schnitzer2001].
This disturbance regime is affected by the current changing context and anthropic activities [@Sist2015], so clarify community responses to disturbance and the underlying ecological processes is crucial to anticipate the fate of tropical forests [@Chazdon2003a].

Disturbances change both biotic and abiotic environment, through modifications in the fluxes of light, heat and water [@Goulamoussene2017], and of the interactions among species such as competition [@Chesson2000].
Community response has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into the recovery of ecosystem processes and services [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that determined succession, mainly driven by the recruitment of pioneer and heliophilous species, shape post-disturbance trajectories [@Carreno-Rocabado2012;@Poorter2016].
Community trajectories would then rely on deterministic processes and would be predictable according to the disturbance intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate from predictions.
Besides, although the deterministic ecological processes at stake are identified, their relative importance compared to stochasticity remains debated [@Norden2015].
This debate translates into the validation and the extent of the intermediate disturbance hypothesis (IDH).
The IDH assumes the predominance of deterministic processes, and predicts a maximum species diversity after moderate disturbance intensity [@Molino2001;@Kariuki2006a].
Validations of the IDH remain scarce in the long-term and often rely on the analysis of taxonomic richness [@Molino2001]. 
Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. 
More ecological-meaningful analysis would couple richness with species evenness, which reveals the changes in the species abundance distribution, and (ii)species composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines community resilience after disturbance [@Elmqvist2003; @Diaz2005].

@@

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
From post-disturbance trajectories (i) we determined the taxonomic and functional translation post-disturbance succession at community scale. (ii) we validated the importance of deterministic processes in community response to disturbance. We defined th interplay of deterministic and stochastic processes along time, which allowed discussing the IDH. (iii) examined community recovery time.

### 2018-12-05 WholePlotTrajectories.tex
intro

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, determine the cycles of
carbon, water and nutrient, and ensure cultural and human well-being.
The growing demand in forests products together with current global
changes increases the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. In this context, the maintenance of
tropical forest community structure, composition and functioning is
increasingly threatened \citep{Anderson-Teixeira2013}.

Tropical forests are subject to natural disturbance events, specifically
tree fall gaps, determining and maintaining the characteristics of
ecosystems \citep{Schnitzer2001}. Disturbances change both the biotic
and abiotic environment, through modifications of the fluxes of light,
heat and water \citep{Goulamoussene2017}, and of the interactions among
species such as competition \citep{Chesson2000}. Understand the drivers
of community response to disturbance and the underlying ecological
processes is then crucial to anticipate the fate of tropical forests in
the changing context \citep{Chazdon2003a}. Community response has been
largely studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into the
recovery of ecosystem processes and services \citep{Herault2018}.
Regarding community diversity and composition, however, post-disturbance
trajectories are not as thoroughly understood. It is recognized that
determined succession, mainly driven by the recruitment of pioneer and
heliophilous species, shape post-disturbance trajectories
\citep{Carreno-Rocabado2012, Poorter2016}. Community trajectories would
then rely on deterministic processes and would be predictable according
to the disturbance intensity and to the characteristics of the
pre-disturbance community. Yet, observed trajectories in diversity and
composition often deviate from predictions. Besides, although the
deterministic ecological processes at stake are identified, their
relative importance compared to stochasticity remains debated
\citep{Norden2015}. This debate translates into the validation and the
extent of the intermediate disturbance hypothesis (IDH). The IDH assumes
the predominance of deterministic processes, and predicts a maximum
species diversity after moderate disturbance intensity
\citep{Molino2001, Kariuki2006a}. Validations of the IDH remain scarce
in the long-term and often rely on the analysis of taxonomic richness
\citep{Molino2001}. Taxonomic richness alone, however, gives limited or
misleading information on forests recovery and functioning
\citep{Chaudhary2016}. More ecological-meaningful analysis would couple
richness with species evenness, which reveals the changes in the species
abundance distribution, and (ii)species composition, which is crucial
for conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore,
a functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

@@

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) examined
community recovery time.

>>>>

Tropical forests cover large areas worldwide and represent crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, determine the cycles of
carbon, water and nutrient, and ensure cultural and human well-being.
However, the growing demand in forests products together with current
global changes increases the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. This context specifically threatens the
maintenance of community structure, composition and functioning
\citep{Anderson-Teixeira2013}. In tropical forests indeed, tree
communities are shaped and maintained by a natural disturbance regime,
specifically tree fall gaps\citep{Schnitzer2001}. This disturbance
regime is affected by the current changing context and anthropic
activities \citep{Sist2015}, so clarify community responses to
disturbance and the underlying ecological processes is crucial to
anticipate the fate of tropical forests \citep{Chazdon2003a}.

Disturbances change both biotic and abiotic environment, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and of the interactions among species such as
competition \citep{Chesson2000}. Community response has been largely
studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into the
recovery of ecosystem processes and services \citep{Herault2018}.
Regarding community diversity and composition, however, post-disturbance
trajectories are not as thoroughly understood. It is recognized that
determined succession, mainly driven by the recruitment of pioneer and
heliophilous species, shape post-disturbance trajectories
\citep{Carreno-Rocabado2012, Poorter2016}. Community trajectories would
then rely on deterministic processes and would be predictable according
to the disturbance intensity and to the characteristics of the
pre-disturbance community. Yet, observed trajectories in diversity and
composition often deviate from predictions. Besides, although the
deterministic ecological processes at stake are identified, their
relative importance compared to stochasticity remains debated
\citep{Norden2015}. This debate translates into the validation and the
extent of the intermediate disturbance hypothesis (IDH). The IDH assumes
the predominance of deterministic processes, and predicts a maximum
species diversity after moderate disturbance intensity
\citep{Molino2001, Kariuki2006a}. Validations of the IDH remain scarce
in the long-term and often rely on the analysis of taxonomic richness
\citep{Molino2001}. Taxonomic richness alone, however, gives limited or
misleading information on forests recovery and functioning
\citep{Chaudhary2016}. More ecological-meaningful analysis would couple
richness with species evenness, which reveals the changes in the species
abundance distribution, and (ii)species composition, which is crucial
for conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore,
a functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines community resilience after
disturbance \citep{Elmqvist2003, Diaz2005}.

@@

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
From post-disturbance trajectories (i) we determined the taxonomic and
functional translation post-disturbance succession at community scale.
(ii) we validated the importance of deterministic processes in community
response to disturbance. We defined th interplay of deterministic and
stochastic processes along time, which allowed discussing the IDH. (iii)
examined community recovery time.
